Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states.

### 2022-06-16 inst/manuscript/manuscript_microxanox.Rmd
ms revised

---
title: "Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states."
author:
- name: Rainer M Krug
  email: Rainer.Krug@uzh.ch;  Rainer@krugs.de
  affiliation: University of Zürich
  footnote: Corresponding Author
- name: Owen L. Petchey
  email: Owen.Petchey@ieu.uzh.ch
  affiliation: University of Zürich
address:
- code: University of Zürich
  address: Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse
    190, 8057 Zurich
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
keywords: 
  - reproducibility
  - regime shift 
  - stable state
  - ordinary differential equations
abstract: |
  $Microxanox$ is an R package to simulate a three functional group system (cyanobacteria, phototrophic sulfur bacteria, and sulfate-reducing bacteria) with four chemical substrates (phosphorus, oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary differential equations. Simulations can be run individually or over a parameter range, to find stable states.  The model features multiple species per functional group, where the number is only limited by computational constraints.
  The R package is constructed in such a way, that the results contain the input parameter used, so that a saved results can be loaded again and the simulation be repeated.
journal: "SoftwareX"
date: "`r Sys.Date()`"

@@

| **Nr.** | **Code metadata description**                                | **Please fill in this column**                                                 |  
| :-----: | :----------------------------------------------------------- | :----------------------------------------------------------------------------- |  
|   C1    | Current code version                                         | v0.9.0                                                                         |  
|   C2    | Permanent link to code/repository used for this code version | [https://github.com/UZH-PEG/microxanox](https://github.com/UZH-PEG/microxanox) | 
|         |                                                              | [10.5281/zenodo.6624125](https://zenodo.org/record/6624125)                    |
|   C3    | Code Ocean compute capsule                                   |                                                                                |  
|   C4    | Legal Code License                                           | MIT                                                                            |  
|   C5    | Code versioning system used                                  | git                                                                            |  
|   C6    | Software code languages, tools, and services used            | [R](https://cran.r-project.org/index.html)                                     |  
|   C7    | Compilation requirements, operating environments             | [R (>= 4.1.0)](https://cran.r-project.org/index.html)                          |  
|         |                                                              | magrittr                                                                       |  
|         |                                                              | tibble                                                                         |  
|         |                                                              | ggplot2                                                                        |  
|         |                                                              | patchwork                                                                      |  
|         |                                                              | grDevices                                                                      |  
|         |                                                              | stats                                                                          |  
|         |                                                              | mgcv                                                                           |  
|         |                                                              | deSolve                                                                        |  
|         |                                                              | dplyr                                                                          |  
|         |                                                              | tidyr                                                                          |  
|         |                                                              | stringr                                                                        |  
|         |                                                              | multidplyr                                                                     |  
|   C8    | If available Link to developer documentation/manual          | [Usert Manual](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html) |  
|   C9    | Support email for questions                                  | Rainer.Krug@uzh.ch; Owen.Petchey@ieu.uzh                                       |

@@

Mathematical models play a key role in the development of understanding about how ecosystems work and how they respond to environmental changes [@Binzer2016a; @Chaparro-Pedraza2021; @Vasseur2005]. They are also important for developing hypotheses that can then be tested in empirical studies. One area of ecology in which simple models have played a influential role is the area of how ecosystems respond to gradual change in an environmental driver [@Scheffer2001]. An environmental driver is a environmental condition that affects an ecosystem, but is assumed to not be affected by the ecosystem, such as the rate of nutrient input into a lake.

@@

An example where a gradual change of an environmental variable causes an abrupt change of the system is the switch from an aerobic (oxygen is available for metabolism) to anaerobic (oxygen generally unavailable) state in a microbial ecosystem. Three types of microbes occur in the model: cyanobacteria (CB) dominating the oxic state, and two types of sulfur bacteria that dominate the anoxic state (sulfate reducing bacteria (SB) and phototrophic sulfur bacteria (PB)). This system has been investigated by @Bush2017 in a simulation study of a mathematical model. They showed that gradual change in the rate at which oxygen could diffuse into the ecosystem (termed the oxygen diffusitivity) could cause catastrophic changes in the ecosystem state that would be difficult to reverse.

One feature of the study by @Bush2017 was limited biodiversity. Specifically, there was no biodiversity within each of the three types of bacteria. This leaves open the question of if and how biodiversity within these types (i.e. functional groups) of bacteria, in their model, affects the ecosystem response to environmental change. This limitation is not specific to the study of @Bush2017, there are few if any studies of the effects of biodiversity on abrupt transitions between ecosystem states.

@@

The second stage was to add functionality that would be necessary to answer our research question. Most importantly, we made it possible to have multiple species of bacteria within each of the three functional groups, for the multiple species to differ in their characteristics, and to vary the number of species and amount of variability among them. We also added functionality that allowed: temporally varying environmental conditions, addition of random noise to state variables, and immigration. In addition to the model itself, the package includes some functions to analyse the results as well as visualize the results to provide a starting point for customized visualizations based on own requirements. The basic and additional functionality is described in the package [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html).

@@

To make the simulation run with multiple species, we expressed different species characteristics in the elements of vectors and matrices. We also coded the ordinary differential equations to include the vectors and matrices, and to use matrix mathematics. In this way, we made it possible to run simulations with different numbers of species without having to change the underlying code.

The package functions and code have modular structure, so that new functionality can be easily added. E.g. temporally defined events of any type could be specified. Further, all parameter values required to run a simulation are stored in one object. Lastly, the general structure of the code should make it straightforward to adapt the model to other similar systems (described in more detail in the Impact section).

@@

The framework used when writing this package aimed to maximise simplicity for the user, and to make it easy to reproduce results (see the supplement [@Petchey2022] to @Limberger2022 for an example of how this is used). As such, all the parameters needed to run a simulation or find a stable state (i.e. the final state of the ecosystem) are contained in a single object (which can easily be created using included functions). This parameter object is given to a function that runs the simulations and returns the results. The returned results object is identical to the parameter object but with an additional slot named `results`, which contains the results of the run. Thus the returned results object contains the simulation conditions (parameters) as well as the results, and can be used to run the simulation again. This promotes reproducibility and makes incremental changes of individual parameters and re-running the simulations straightforward.

@@

Two methods for finding steady states are implemented. The first runs a separate simulation for each combination of starting conditions and oxygen diffusivity (we term this the *Replication method*). This is the method used in the @Bush2017 study. The second runs two simulations, with step-wise and slowly temporally increasing oxygen diffusivity, and the other with step-wise and slowly decreasing or decreasing oxygen diffusivity. During this temporal environmental change, the state of the system was recorded just before change to a new oxygen diffusivity (we term this the *Temporal method*). We implemented two methods since there is no definitive best method, and in order to check if results were sensitive to choice of method.

The replication method is implemented in the function `run_replication_ssfind()` which takes a parameter object as returned by the function `new_replication_ssfind_parameter()` and the number of cores for multithreading the simulation. As the multithreading uses the package function `mclapply()` from the package `parallel` [@RCoreTeam2022], the multthreading only works on Linux and Mac. It is planned to move to `parLapply()` [@RCoreTeam2022] in a future release.

This temporal method implemented in the function `run_temporal_ssfind()`, which takes a parameter object as created by the function `new_temporal_ssfind_parameter()`. It is planned for a later release, to run these two simulations in parallel.

@@

Using the package to calculate the extent of hysteresis involves setting ecosystem and simulation parameters, including parameters for the finding of stable states across an environmental gradient, running the stable state finding function, and analysing the results with the function that calculates extent of hysteresis. The code for this is provided in the *User Guide* vignette.

@@

Researchers wanting to model a new ecosystem do not, therefore, have to start from scratch. This will relieve researchers from needing to making software design decisions, and rather focus on appropriately representing their ecosystem, and finding the results that interest them. Nevertheless, adaptation of the code in the package will require a person / persons that can take a conceptual model of an ecosystem and then represent that in terms of parameters and rate equations, and that is relatively proficient in R programming.

>>>>

---
title: "Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states."
author:
- name: Rainer M Krug
  email: Rainer.Krug@uzh.ch;  Rainer@krugs.de
  affiliation: University of Zürich
  footnote: Corresponding Author
- name: Owen L. Petchey
  email: Owen.Petchey@ieu.uzh.ch
  affiliation: University of Zürich
address:
- code: University of Zürich
  address: Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse
    190, 8057 Zurich
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
keywords: 
  - reproducibility
  - regime shift 
  - stable state
  - ordinary differential equations
abstract: |
  $Microxanox$ is an R package to simulate a three functional group ecosystem (cyanobacteria, phototrophic sulfur bacteria, and sulfate-reducing bacteria) with four chemical substrates (phosphorus, oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary differential equations. Simulations can be run individually or over a parameter range, to find stable states.  The model can be implemented with different numbers of species per functional group. The package is constructed in such a way that the results contain the input parameter used, so that a saved results can be loaded again and the simulation be repeated. Furthermore, the package framework and code should serve as a useful starting point for making simulation models of other types of ecosystem.
journal: "SoftwareX"
date: "`r Sys.Date()`"

@@

| **Nr.** | **Code metadata description**                                | **Please fill in this column**                                                 |  
| :-----: | :----------------------------------------------------------- | :----------------------------------------------------------------------------- |  
|   C1    | Current code version                                         | v0.9.0                                                                         |  
|   C2    | Permanent link to code/repository used for this code version | [https://github.com/UZH-PEG/microxanox](https://github.com/UZH-PEG/microxanox) | 
|         |                                                              | [10.5281/zenodo.6624125](https://zenodo.org/record/6624125)                    |
|   C3    | Code Ocean compute capsule                                   |                                                                                |  
|   C4    | Legal Code License                                           | MIT                                                                            |  
|   C5    | Code versioning system used                                  | git                                                                            |  
|   C6    | Software code languages, tools, and services used            | [R](https://cran.r-project.org/index.html)                                     |  
|   C7    | Compilation requirements, operating environments             | [R (>= 4.1.0)](https://cran.r-project.org/index.html)                          |  
|         |                                                              | magrittr                                                                       |  
|         |                                                              | tibble                                                                         |  
|         |                                                              | ggplot2                                                                        |  
|         |                                                              | patchwork                                                                      |  
|         |                                                              | grDevices                                                                      |  
|         |                                                              | stats                                                                          |  
|         |                                                              | mgcv                                                                           |  
|         |                                                              | deSolve                                                                        |  
|         |                                                              | dplyr                                                                          |  
|         |                                                              | tidyr                                                                          |  
|         |                                                              | stringr                                                                        |  
|         |                                                              | multidplyr                                                                     |  
|   C8    | If available Link to developer documentation/manual          | [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html) |  
|   C9    | Support email for questions                                  | Rainer.Krug@uzh.ch; Owen.Petchey@ieu.uzh                                       |

@@

Mathematical models play a key role in the development of understanding about how ecosystems work and how they respond to environmental changes [@Binzer2016a; @Chaparro-Pedraza2021; @Vasseur2005]. They are also important for developing hypotheses to test in empirical studies. One area of ecology in which models have played a influential role is how ecosystems respond to gradual change in an environmental driver [@Scheffer2001]. An environmental driver is a environmental condition that affects an ecosystem, but is assumed to not be affected by the ecosystem, such as the rate of nutrient input into a lake.

@@

An example where a gradual change of an environmental variable causes an abrupt change of the system is the switch from an aerobic (oxygen is available for metabolism) to anaerobic (oxygen generally unavailable) state in a microbial ecosystem. This system has been investigated by @Bush2017 in a simulation study of a mathematical model. Three types of microbes occur in the model: cyanobacteria (CB) dominating the oxic state, and two types of sulfur bacteria that dominate the anoxic state (sulfate reducing bacteria (SB) and phototrophic sulfur bacteria (PB)). The model shows that gradual change in the rate at which oxygen could diffuse into the ecosystem (termed the oxygen diffusitivity) could cause catastrophic changes in the ecosystem state that would be difficult to reverse.

One feature of the study by @Bush2017 was limited biodiversity. Specifically, there was no biodiversity within each of the three types of bacteria. This leaves open the question of if and how biodiversity within these types (i.e. functional groups) of bacteria affects the ecosystem response to environmental change. This limitation is not specific to the study of @Bush2017. There are few if any studies of the effects of biodiversity on abrupt transitions between ecosystem states.

@@

The second stage was to add functionality that would be necessary to answer our research question. Most importantly, we made it possible to have multiple species of bacteria within each of the three functional groups, for the multiple species to differ in their characteristics, and to vary the number of species and amount of variability among them. We also added functionality that allowed: temporally varying environmental conditions, addition of random noise to state variables, and immigration. In addition to the model itself, the package includes some functions to analyse the results as well as to visualize the results to provide a starting point for customized visualizations based on own requirements. The basic and additional functionality is described in the package [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html).

@@

To make the simulation run with multiple species per functional group, we expressed different species characteristics in the elements of vectors and matrices. We also coded the ordinary differential equations to include the vectors and matrices, and to use matrix mathematics. In this way, we made it possible to run simulations with different numbers of species without having to change the underlying code.

The package functions and code have modular structure so that new functionality can be easily added. E.g. temporally defined events of any type could be specified. Further, all parameter values required to run a simulation are stored in one object. Lastly, the general structure of the code should make it straightforward to adapt the model to other similar systems (described in more detail in the Impact section).

@@

The framework used when writing this package aimed to maximise simplicity for the user, and to make it straightforward to reproduce results (see the supplement [@Petchey2022] to @Limberger2022 for an example of how this is used). As such, all the parameters needed to run a simulation or find a stable state (i.e. the final state of the ecosystem) are contained in a single object (which can easily be created using included functions). This parameter object is given to a function that runs the simulations and returns the results. The returned results object is identical to the parameter object but with an additional slot named `results`, which contains the results of the run. Thus the returned results object contains the simulation conditions (parameters) as well as the results, and can be used to run the simulation again. This promotes reproducibility and makes incremental changes of individual parameters and re-running the simulations straightforward.

@@

Two methods for finding steady states are implemented. The first runs a separate simulation for each combination of starting conditions and oxygen diffusivity (we term this the *Replication method*). This is the method used in the @Bush2017 study. The second runs two simulations, one with step-wise and slowly temporally *increasing* oxygen diffusivity, and the other with step-wise and slowly *decreasing* oxygen diffusivity. During this temporal environmental change, the state of the system is recorded just before change to a new oxygen diffusivity (we term this the *Temporal method*). We implemented two methods since there is no definitive best method, and in order to check if results were sensitive to choice of method.

The replication method is implemented in the function `run_replication_ssfind()` which takes a parameter object as returned by the function `new_replication_ssfind_parameter()` and the number of cores for multithreading the simulation. As the multithreading uses the package function `mclapply()` from the package `parallel` [@RCoreTeam2022], the multithreading only works on Linux and Mac. It is planned to move to `parLapply()` [@RCoreTeam2022] in a future release.

The temporal method is implemented in the function `run_temporal_ssfind()`, which takes a parameter object as created by the function `new_temporal_ssfind_parameter()`. It is planned for a later release, to run these two simulations in parallel.

@@

Using the package to calculate the extent of hysteresis involves setting ecosystem and simulation parameters, including parameters for the finding of stable states across an environmental gradient, running the stable state finding function, and analysing the results with the function that calculates extent of hysteresis. The code for this is provided in the *User Guide*.

@@

Researchers wanting to model a new ecosystem do not, therefore, have to start from scratch. This will relieve researchers from needing to make software design decisions, and rather focus on appropriately representing their ecosystem and finding the results that interest them. Nevertheless, adaptation of the code in the package will require a person / persons that can take a conceptual model of an ecosystem and then represent that in terms of parameters and rate equations, and that is relatively proficient in R programming.
